Estimating Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks

### 2024-07-02 paper/main.tex
Fixes after SZ review

\documentclass{article}

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=apa, maxcitenames=1, uniquelist=false, url=false]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}
\usepackage[colorlinks=true, allcolors=blue, breaklinks=true]{hyperref}
\usepackage{xurl}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{color}
\usepackage{tabularray}

\defbibenvironment{bibliography}
  {\begin{enumerate}}
  {\end{enumerate}}
  {\item}
\AtEveryBibitem{%
  \clearfield{note}%
}

% \AtEveryCitekey{\iffootnote{\csxdef{blx@footnotemark}{}\csxdef{blx@shortfootnotemark}{}}{}}

% \DeclareCiteCommand{\cite}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\usebibmacro{cite}}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

%   \DeclareCiteCommand{\parencite}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\bibopenparen\usebibmacro{cite}\bibcloseparen}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

% % Redefine \parencites to handle multiple citations in one set of parentheses
% \DeclareCiteCommand{\parencites}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\bibopenparen\usebibmacro{cite}\bibcloseparen}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

@@

\title{Predicting Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks}
\author[1,*]{Paweł Antoniuk}
\author[1]{Sławomir Zieliński}
\affil[1]{Faculty of Computer Science, Białystok University of Technology}
\affil[*]{Corresponding author: pawel.antoniuk@sd.pb.edu.pl}
\date{}

\begin{document}
\maketitle

\begin{abstract}
  Binaural audio technology has been in existence for many years, but its popularity has significantly increased over the past decade as a consequence of advancements in virtual reality and streaming technologies. Along with its growing popularity, the quantity of publicly accessible binaural audio materials has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble location and width being the most prominent. This study presents a novel method for predicting ensemble location and width in binaural recordings of music. To this end, 30 head-related transfer functions and 192 binaural music recordings from publicly accessible multi-track recording repositories were used to synthesize 23,040 binaural recordings. The synthesized recordings were then used to train a multi-task convolutional neural network model, the aim of which was to predict the location and width of the ensemble for unseen recordings. The results indicate that models of ensemble location and width can be successfully constructed with low prediction errors --- $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. The method developed in this study outperforms previous spatiogram-based methods recently published in the literature and holds promise for future development as a part of a novel tool for audio engineers to assess binaural audio.
\end{abstract}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This is possible by internal examination of interaural differences in time, loudness, and frequency, known as binaural hearing \parencite{blauert_spatial_1996}, which enables precise localization of sound sources in complex auditory environments. A notable advantage of binaural hearing is exemplified by the `cocktail party effect', highlighting the system's capacity to concentrate on foreground sound sources while suppressing background noise \parencite{cherry_experiments_1953}.

Understanding of the auditory system is essential for comprehending its limits and creating more immersive binaural experiences for entertainment purposes \parencite{zhang_surround_2017}. It is also important for enhancing auditory signal reception in hearing aid devices \parencite{hirsh_binaural_1950, thiemann_speech_2016}.

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has prompted an intriguing exploration into the extent to which these tools can emulate the human auditory system. This exploration proceeds without reliance on the advanced spatial audio feature engineering traditionally used in audio source localization, as evidenced by some prominent recent examples \parencite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. While the application of convolutional neural networks (CNNs) \parencite{lecun_handwritten_1989} --- traditionally employed in the visual domain --- to audio signals, particularly in conjunction with spectrograms \parencite{thomas_analyzing_2014, espi_exploiting_2015, han_convolutional_2017} or other feature engineering techniques \parencite{abdel-hamid_applying_2012, sainath_deep_2013}, is not novel. However, these approaches are continually being refined and adapted to the audio realm. Building on these foundations, this study developed an audio localization method by constructing a spectrogram-based multi-task CNN model.

Inspired by the fact, that humans tend to localize sound sources in groups rather than individually \parencite{bregman_auditory_1990, rumsey_spatial_2002}, the objective of the proposed model is to predict ensemble location and width instead of positions of individual sources. This study is unique as it not only conceptualized the method but also tested it on a vast, real-life music corpus of 23,040 binaural excerpts that were synthesized using 192 multi-track music recordings \parencite{senior_mixing_2023} and 30 publicly available head-related transfer functions (HRTFs) from various sources (see Table \ref{table:hrtfs} in \ref{appendix:a_hrtf}). The music recordings covered many different types of genres, including rock, jazz, pop, and classical music.

The findings demonstrate that this method is effective in accurately predicting the spatial characteristics of sound sources in near-real-world scenarios. Furthermore, this paper demonstrates an experimental framework that facilitates the objective measurement of a binaural localization technique, employing a large-scale dataset synthesized from real-world music signal (for applications of similar frameworks, see: \cite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}). One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, significant limitations of this study include the absence of reverberation in the synthesized recordings and the method's inapplicability to real-time scenarios --- both are critical areas for future research.

The developed method has the potential to be highly beneficial in automated assessment tasks, where a significant number of binaural recordings must be measured or labeled in terms of their spatial content information. This could be utilized in the development of a hypothetical autonomous web-crawler bot that will collect binaural recordings from publicly accessible repositories and label them according to the spatial properties of the sound sources, such as the location of the music ensemble or the sparsity of audio source positions. This method may also assist audio engineers in objectively assessing and segregating binaural audio recordings with regard to their spatial content.

The structure of this paper is organized as follows. Section \ref{sec:related-studies} presents related studies. The description of the method developed for this study is provided in Section \ref{sec:methodology}, which also includes detailed definitions of ensemble location and width, along with a description of the experiments used to evaluate this method. Section \ref{sec:results} presents and discusses the performance of the presented method and the results of the experiments conducted in this study. The paper concludes in Section \ref{sec:conclusions}.

@@

The majority of existing literature on the subject of sound source localization employs techniques that leverage the advantages of microphone arrays with more than two channels \parencite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. These method simplify the task and can improve the localization performance; however, they do not utilize binaural hearing, rendering them ineffective for binaural recordings.

In the context of sound source localization in binaural signals, the  focus of research is on the identification of individual sound sources, rather than groups of sounds \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}. In terms of source direction of arrival (DOA) methods, the majority of research assumes a fixed number of sound sources \parencite{pang_multitask_2019, vera-diaz_towards_2018, ma_exploiting_2017, woodruff_binaural_2012, arthi_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Furthermore, the approach is not limited to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to previous studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as a group, or in this case, a musical ensemble. This approach is similar to how the real musical ensembles are arranged on stage, thereby emphasizing practical applications in live settings. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk_blind_2023} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

Sound localization methods can be classified into two categories based on the implementation of their underlying algorithms: glass-box (e.g., \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, may_robust_2015, woodruff_binaural_2012, ma16c_interspeech, ma_exploiting_2017, ma_robust_2018}) and black-box (e.g., \parencite{vera-diaz_towards_2018, vera-diaz_towards_2018, yang_deepear_2022}). Glass-box methods, more traditional in the literature, rely on manually designed algorithms that mimic the auditory system to explicitly extract key features for location prediction, such as interaural level differences, interaural time differences, interaural coherence, or interaural phase differences (see \cite{blauert_spatial_1996} for feature descriptions). An example of one of the most advanced auditory models that is able to extract such features was developed by the Two!Ears project \parencite{raake_computational_nodate}.

In contrast, black-box methods generally employ minimal feature engineering, relying instead on advanced machine learning techniques to autonomously extract features and make predictions. These methods, while effective, do not explicitly reveal the features they extract nor necessarily mimic the human auditory system. This opacity and the unpredictable nature of the outcomes, coupled with their dependency on deep neural networks that have extensive learning parameters, require the use of very large datasets for development and evaluation. Such datasets often include thousands of examples, as seen in the TIMIT corpus with 6300 examples \parencite{garofolo_darpa_1993} used in multiple studies \parencite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vera-diaz_towards_2018, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}, or custom corpora comprising hundreds of thousands of recordings \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}. This poses a significant challenge in assembling a sufficiently large and diverse collection of labeled binaural recordings. However, this challenge can be addressed through the synthesis of binaural sounds, as demonstrated in various studies \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020,yang_deepear_2022, ma_robust_2018} and further elaborated upon in Section \ref{subsec:synthesis} of this paper.

@@

This section presents a detailed description of the main objective of the model developed as part of this study, as outlined in Section \ref{subsec:ensemble_definition}. It also describes the audio dataset used for training and evaluating the model, as detailed in Section \ref{subsec:synthesis}. In Section \ref{subsec:feature_extraction}, the feature extraction procedure is presented. Section \ref{subsec:topology} describes the model topology, whereas Section \ref{subsec:training_evaluation} address model training and evaluation.

@@

The objective of the model developed in this study is to predict the ensemble location ($\theta$) and width ($\omega$), as illustrated in Figure \ref{fig:scene}. An ensemble is defined as a group of audio point sources positioned on a circle around the listener on a virtual acoustic scene with equal distance to the listener. The location of source $i$ is denoted by $\theta_i$. The ensemble width ($\omega$) is defined as the angular width between two extreme point sources ($max_i(\theta_i)-min_i(\theta_i)$), while the ensemble location, designated by $\theta$, represents the middle angle between two extreme sound sources ($(max_i(\theta_i)+min_i(\theta_i))/2$). For the purposes of this study, the locations of the sources were limited to the frontal hemisphere only, i.e. $\theta\in[-45\degree, 45\degree]$, $\omega\in[0\degree, 90\degree]$, as this range encompasses the majority of real-world recording scenarios. It should be noted that although humans possess some limited abilities to localize sound sources in the vertical plane, in this study all sources are placed in the horizontal plane, at the height of the listener. This covers the majority of cases for real-world recordings (see \cite{ma_robust_2018, zielinski_spatial_2022} for related studies that cover top-down discrimination).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{../pictures/scene.png}
  \caption{\label{fig:scene}Illustration of ensemble width ($\omega$) and ensemble location ($\theta$) that are relative to the direction of the head. Black dots represent the positions of audio sources $\theta_i$. The ensemble location ($\theta$) is defined as the angular position of the center of the ensemble relative to the direction the head is turned. The ensemble width ($\omega$) is defined as angular distance between two extreme audio sources. }
\end{figure}

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. The binaural recordings were synthesized using 192 multi-track publicly-available music recordings \parencite{senior_mixing_2023} and 30 HRTF databases (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list of HRTF databases used in this study). The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora. Before the synthesis, the tracks were equalized to $-23$ LUFS, in accordance with the \parencite{noauthor_itu-r_2023} recommendation.

@@

Due to copyright restrictions, the music corpus utilized in this study was not published and can be provided upon request to the authors of this paper.

\subsection{Feature extraction}
\label{subsec:feature_extraction}

% The procedure proposed in this study did not 
% Zamiast tego dac obrazek z glowami i spektrogramami, lewy i prawi i roznica, sprawdzic czy to ma sens, potencjalnie usunac
% \begin{figure}[ht]
%   \centering
%   \includegraphics[width=\linewidth]{../figures/samples.pdf}
%   \caption{\label{fig:samples}Example magnitude spectrograms from randomly selected synthesized binaural music recordings, used as input data for training and evaluating the convolutional neural network (CNN) in this study. }
% \end{figure}
% Figure \ref{fig:samples} displays example spectrograms from four distinct binaural recordings, exhibiting a broad bandwidth and the diversity of the signals used in this study

Prior to input into the model, the binaural recordings of music were transformed into magnitude spectrograms. Although spectrograms do not directly provide information that can be translated into ensemble features, especially ensemble width, the goal of this task was to reduce the number of independent variables compared to the raw audio signal by extracting more compressed and informative data in the frequency domain. This step was necessary to decrease the likelihood of overfitting, reduce the number of examples needed to train the model, and thereby lower the overall computational power requirements. It is worth mentioning, however, that recent methods have shown that CNNs can be suitable for end-to-end audio localization without the spectrogram extraction step \parencite{vera-diaz_towards_2018, vecchiotti_end--end_2019}.

To prepare the input for the model, a Hamming window of 40 ms was applied to each frame of the signal, resulting in a total of 349 time frames. From each frame, spectrograms were extracted using the Fast Fourier Transform (FFT) algorithm, with 150 frequency bands spaced linearly from 100 Hz to 16 kHz. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Consequently, each sample was represented by a floating-point precision matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented by \cite{zielinski_automatic_2022}.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{../pictures/architecture.pdf}
  \caption{\label{fig:architecture}Topology of the Convolutional Neural Network (CNN) used for identifying ensemble location and width}
\end{figure}

The network topology employed in this study was heavily influenced by the AlexNet convolutional neural network introduced by \cite{krizhevsky_imagenet_2012}. While AlexNet was originally designed for image classification, it was adapted for the audio prediction task in this study by converting binaural recordings into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. This conversion allowed the spectrograms to be treated like images, allowing them to be used in a standard image-recognition-like task.

As illustrated in Figure \ref{fig:architecture}, the network architecture consists of an input layer accepting a pair of spectrograms, followed by a series of convolutional units and classification units, culminating in two output heads for predicting ensemble width and location. This design employs a multi-task approach, enabling a single network to predict both ensemble parameters simultaneously.

The topology finalized in this study was constructed by selecting from many alternative architectures, choosing the one with the highest prediction quality observed on the validation dataset. Despite the existence of numerous algorithms for automatic topology selection \parencite{branke_evolutionary_1995, zhang_finding_2018, miikkulainen_evolving_2017, stanley_evolving_2002, shafiee_deep_2016}, the final topology was determined manually, primarily due to the high computational demands relative to the available resources.

>>>>

\documentclass[11pt]{article}

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=2cm,right=2cm,marginparwidth=1cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=custom, maxcitenames=1, uniquelist=false, uniquename=false, url=false]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}
\usepackage[colorlinks=true, allcolors=blue, breaklinks=true]{hyperref}
\usepackage{xurl}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{color}
\usepackage{tabularray}
\usepackage{indentfirst}
\usepackage{setspace}
\usepackage{orcidlink}

\doublespacing
\defbibenvironment{bibliography}
  {\begin{enumerate}}
  {\end{enumerate}}
  {\item}
\AtEveryBibitem{%
  \clearfield{note}%
}
\renewcommand\Affilfont{\itshape\centering}
\renewenvironment{abstract}{}{}
\newcommand{\keywords}[1]{
  \noindent\begin{center}\begin{minipage}{\textwidth}
  \small\textbf{Keywords:} #1
  \end{minipage}\end{center}
}
\DefineBibliographyStrings{english}{%
  andothers = {\textit{et al\adddot}},
}
\DeclareFieldFormat{doi}{%
  \href{https://doi.org/#1}{#1}%
}
\renewbibmacro*{doi+eprint+url}{%
  \iftoggle{bbx:doi}
    {\printfield{doi}}
    {}%
  \newunit
  \iftoggle{bbx:eprint}
    {\usebibmacro{eprint}}
    {}%
  \newunit
  \iftoggle{bbx:url}
    {\usebibmacro{url+urldate}}
    {}}

@@

\title{Estimating Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks}
\author[1,*]{Paweł Antoniuk\orcidlink{0000-0002-0914-8920}}
\author[1]{Sławomir Zieliński\orcidlink{0000-0002-3205-974X}}
\affil[1]{Faculty of Computer Science, Białystok University of Technology \protect\\ Wiejska 45A, 15-351 Białystok, Poland}
\affil[*]{Corresponding Author Email: pawel.antoniuk@sd.pb.edu.pl}
\date{}

\begin{document}

\maketitle

\begin{abstract}
  Binaural audio technology has been in existence for many years. However, its popularity has significantly increased over the past decade as a consequence of advancements in virtual reality and streaming techniques. Along with its growing popularity, the quantity of publicly accessible binaural audio recordings has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble location and width being the most prominent. This study presents a novel method for estimating these ensemble parameters in binaural recordings of music. For this purpose, a dataset of 23,040 binaural recordings was synthesized from 192 publicly-available music recordings using 30 head-related transfer functions. The synthesized excerpts were then used to train a multi-task spectrogram-based convolutional neural network model, aiming to estimate the ensemble location and width for unseen recordings. The results indicate that a model for estimating ensemble parameters can be successfully constructed with low prediction errors: $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. The method developed in this study outperforms previous spatiogram-based techniques recently published in the literature and shows promise for future development as part of a novel tool for binaural audio recordings analysis.
\end{abstract}

\keywords{ensemble width, ensemble location, binaural, spatial audio, localization, convolutional neural network, head-related transfer function, angle of arrival}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This is possible, among other factors, by internal examination of interaural differences in time, loudness, and frequency, known as binaural hearing \parencite{blauert_spatial_1996}, which enables precise localization of sound sources in complex auditory environments. A notable advantage of binaural hearing is exemplified by the ``cocktail party effect'', highlighting humans' capability to concentrate on foreground sound sources while suppressing background noise \parencite{cherry_experiments_1953}. Understanding of the auditory system is essential for comprehending its limits but also for leveraging these insights to create more immersive binaural experiences for entertainment purposes \parencite{zhang_surround_2017}. It is also important for enhancing auditory signal reception in hearing aid devices \parencite{hirsh_binaural_1950, thiemann_speech_2016}.

The advance of sophisticated machine learning techniques, especially deep learning networks, has initiated an interesting exploration of their potential to emulate the human auditory system. Recently emerged studies have demonstrated that relying on the advanced spatial audio feature engineering is not necessary in computational audio source localization \parencite{vera-diaz_towards_2018, pang_multitask_2019, yang_deepear_2022}. While applying convolutional neural networks (CNNs) \parencite{lecun_handwritten_1989} to audio signals is well-established, often in conjunction with spectrograms \parencite{thomas_analyzing_2014, espi_exploiting_2015, han_convolutional_2017} or other feature engineering techniques \parencite{abdel-hamid_applying_2012, sainath_deep_2013}, these approaches continue to be refined and adapted for audio processing. Building on these foundations, this study develops an audio localization method using a spectrogram-based multi-task CNN model.

Humans tend to localize groups of sound sources rather than individual ones \parencite{bregman_auditory_1990, rumsey_spatial_2002}. Inspired by this fact, the objective of the proposed model is to estimate the location and width of these groups, termed ensembles, instead of the positions of individual sources. This study is unique as it not only developed the method but also tested it on a vast, realistic music corpus. The corpus comprised 23,040 binaural excerpts synthesized using 192 multi-track music recordings (from a repository provided by \textcite{senior_mixing_2023}) and 30 publicly available head-related transfer functions (HRTFs) from various sources (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list). The music recordings covered many different genres, including rock, jazz, pop, and classical music.

The findings demonstrate that this method is effective in accurately estimating the spatial characteristics of groups of sound sources in near-real-world scenarios. This paper also demonstrates an experimental framework that facilitates the objective measurement of a binaural localization technique, employing a large-scale dataset synthesized from real-world music signals (for applications of similar frameworks, see studies conducted by \textcite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}). One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, significant limitations of this study include the absence of reverberation in the synthesized recordings and the method's inapplicability to real-time scenarios --- both are critical areas for future research.

The developed method has the potential to be highly beneficial in automated assessment tasks, where a significant number of binaural recordings must be measured or labeled in terms of their spatial content information. This could be utilized in the development of a hypothetical autonomous ``web-crawler bot'' that will collect binaural recordings from publicly accessible repositories and label them according to the spatial properties of the sound sources, such as the location of the music ensemble or the sparsity of audio source positions. This method may also assist audio engineers in objectively assessing and segregating binaural audio recordings with regard to their spatial content.

This paper is structured as follows: Section \ref{sec:related-studies} presents related studies. The description of the method developed for this study is provided in Section \ref{sec:methodology}, which also includes detailed definitions of ensemble location and width, along with a description of the experiments used to evaluate this method. Section \ref{sec:results} presents and discusses the performance of the proposed method as well as the results of the experiments conducted in this study. Finally, Section \ref{sec:conclusions} offers concluding remarks and suggestions for future research.

@@

Most existing literature on computational sound source localization reports techniques that take advantage of multiple microphone arrays with more than two channels \parencite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. Although these methods can improve localization precision by providing additional spatial information, they do not utilize binaural hearing, rendering them ineffective for binaural recordings. In the context of sound source localization in binaural signals, the  focus of research is put on the identification of individual sound sources, rather than groups of sounds \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}.

Considering source Direction of Arrival (DoA) methods, the majority of research assumes a fixed number of sound sources \parencite{pang_multitask_2019, vera-diaz_towards_2018, ma_exploiting_2017, woodruff_binaural_2012, arthi_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Moreover, the approach is not narrowed to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as a group, or in this case --- a musical ensemble --- similar to how real musical ensembles are arranged on stage. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \textcite{antoniuk_blind_2023} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

Sound localization methods can be classified into two categories based on the implementation of their underlying algorithms, termed as glass-box and black-box techniques. Glass-box methods could be considered as more traditional in the literature. They rely on manually designed algorithms that mimic the auditory system to explicitly extract key features for the localization estimation, such as interaural level differences, interaural time differences, interaural coherence, or interaural phase differences (\textcite{blauert_spatial_1996} provides detailed descriptions of these features). Examples of glass-box methods can be found in numerous studies, including those conducted by \textcite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, may_robust_2015, woodruff_binaural_2012, ma16c_interspeech, ma_exploiting_2017, ma_robust_2018, zielinski_automatic_2022}. These features are typically extracted using an auditory model. An advanced implementation capable of extracting these features was developed as part of the Two!Ears project \parencite{raake_computational_2016}.

Black-box methods use a minimal degree of feature engineering, depending on deep neural networks to both extract features and make estimations. While effective, these methods do not necessarily consistently mimic human hearing, rendering them less suitable for objective measurement tasks (e.g., \textcite{vera-diaz_towards_2018, yang_deepear_2022}). Additionally, it is challenging to reveal their internally extracted features. Due to their opacity, unpredictable results, and numerous learning parameters, these methods should be treated more carefully. Moreover, they require large datasets for their development and evaluation. These datasets often contain thousands of examples, such as the TIMIT corpus \parencite{garofolo_darpa_1993} used in multiple studies \parencite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vera-diaz_towards_2018, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}. Some researchers have even created custom corpora with hundreds of thousands of recordings \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}.

The necessity of having a large corpus to train deep learning models poses a significant challenge in gathering a sufficiently large and diverse collection of labeled binaural recordings. However, this challenge can be addressed through the synthesis of binaural sounds, as demonstrated in various studies \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020, yang_deepear_2022, ma_robust_2018} and discussed further in Section \ref{subsec:synthesis}.

@@

This part of the paper presents a detailed description of the model developed in this study, as outlined in Section \ref{subsec:ensemble_definition}. It also describes the audio dataset used for training and evaluating the model, as detailed in Section \ref{subsec:synthesis}. In Section \ref{subsec:feature_extraction}, the spectrograms calculation procedure is presented. Section \ref{subsec:topology} describes the model topology, whereas Section \ref{subsec:training_evaluation} addresses model training and evaluation.

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=\linewidth]{../pictures/scene2.pdf}
  \caption{\label{fig:scene}Illustration of ensemble width ($\omega$) and ensemble location ($\theta$) relative to the direction of the head orientation. Black dots represent the positions of audio sources $\theta_i$. The ensemble location ($\theta$) is the angular position of the center of the ensemble relative to the direction the head is facing. The ensemble width ($\omega$) is the angular distance between the two most extreme audio sources in the ensemble.}
\end{figure}

The objective of the model developed in this study is to estimate the ensemble location ($\theta$) and width ($\omega$), as illustrated in Figure \ref{fig:scene}. An ensemble is defined as a group of audio point sources positioned on a circle around the listener on a virtual acoustic scene with equal distance to the listener. The location of source $i$ is denoted by $\theta_i$. The ensemble width ($\omega$) is defined as the angular distance between two extreme point sources ($max_i(\theta_i)-min_i(\theta_i)$), while the ensemble location, designated by $\theta$, represents the middle angle between two extreme sound sources ($(max_i(\theta_i)+min_i(\theta_i))/2$). For the purposes of this study, the locations of the sources were limited to the frontal hemisphere only, i.e. $\theta\in[-45\degree, 45\degree]$, $\omega\in[0\degree, 90\degree]$, as this range encompasses the majority of real-world recording scenarios. It should be noted that although humans possess some limited abilities to localize sound sources in the vertical plane, in this study all sources are placed in the horizontal plane, at the ear-level of the listener. This covers the majority of cases for real-world recordings (see \textcite{ma_robust_2018, zielinski_spatial_2022} for related studies that cover top-down discrimination).

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. These recordings were synthesized using 192 publicly-available multi-track music recordings \parencite{senior_mixing_2023} and 30 HRTF databases (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list). The large number of HRTF databases was necessary to make the model as generalizable as possible. In real-world scenarios, the HRTF used for binaural synthesis is often unknown, so constructing a model for a single HRTF would have limited practical utility. The aim was to predict ensemble parameters regardless of the specific HRTF function used. Additionally, the large number of HRTF functions increased the amount of data available for model training, which is particularly beneficial in the context of deep neural networks. The number of HRTF databases (30) was determined using heuristics from previous study conducted by \textcite{zielinski_automatic_2022}, which suggested this number should be sufficient for the task.

The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora. Before the synthesis, the signals in each track were equalized to $-23$ LUFS, in accordance with the ITU-R BS.1770-5 (\citeyear{noauthor_itu-r_2023}) recommendation.

@@

After the binauralization procedure, the synthesized recordings were truncated to a duration of seven seconds, with sine-squared fade-in and fade-out effects of 0.01 seconds applied. The recordings were then RMS-normalized, scaled by a factor of 0.9, and DC-offset corrected. They were stored as uncompressed files at a 48 kHz sampling rate and with a 32-bit resolution.

Due to copyright restrictions, the music corpus utilized in this study was not published. However, the corpus can be provided upon reasonable request from the authors of this paper.

\subsection{Calculation of spectrograms}
\label{subsec:feature_extraction}

Prior being input into the model, the binaural recordings of music were transformed into magnitude spectrograms. Although spectrograms do not directly provide information that can be translated into ensemble features, especially ensemble width, the goal of this task was to reduce the number of independent variables compared to the raw audio signal by extracting more compressed and informative data in the frequency domain. This step was necessary to decrease the likelihood of overfitting, reduce the number of examples needed to train the model, and thereby lower the overall computational power requirements. It is worth mentioning, however, that recent methods have shown that CNNs are suitable for end-to-end audio localization without the spectrogram extraction step, as shown by \textcite{vera-diaz_towards_2018, vecchiotti_end--end_2019}.

To prepare the input for the model, a Hamming window of 40 ms with an overlap of 20 ms was applied to each frame of the signal, resulting in a total of 349 time frames. From each frame, spectrograms were extracted using the Fast Fourier Transform (FFT) algorithm, with 150 frequency bands spaced linearly from 100 Hz to 16 kHz. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Consequently, each sample was represented by a 32-bit floating-point precision matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented by \textcite{zielinski_automatic_2022}.

@@

The network topology employed in this study was strongly influenced by the AlexNet convolutional neural network introduced by \textcite{krizhevsky_imagenet_2012}. While AlexNet was originally designed for image classification, in this study it was adapted for the audio analysis task by converting binaural recordings into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. This conversion allowed the spectrograms to be treated like visual data, enabling them to be used in an image-recognition-like task.

As illustrated in Figure \ref{fig:architecture}, the network architecture consists of an input layer accepting a pair of spectrograms, followed by a series of convolutional units and classification units, culminating in two outputs predicting ensemble location and width, respectively. This design employs a multi-task approach, enabling a single network to estimate both ensemble parameters simultaneously.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=\linewidth]{../pictures/architecture.pdf}
  \caption{\label{fig:architecture}Topology of the Convolutional Neural Network (CNN) used for estimating ensemble location and width, illustrating the layers (grouped in ``convolutional'' and ``classification'' units) and connections of the network architecture.}
\end{figure}

The topology finalized in this study was chosen, among many alternative architectures, based on the highest prediction quality observed on the validation dataset. Despite the existence of numerous algorithms for automatic topology selection \parencite{branke_evolutionary_1995, zhang_finding_2018, miikkulainen_evolving_2017, stanley_evolving_2002, shafiee_deep_2016}, the final topology was determined manually, primarily due to the high computational demands relative to the available resources.
